---
title: "Getting Started with tidybari"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with tidybari}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE  # Most chunks require network access to Dataverse
)
```

## Introduction

The **tidybari** package provides easy access to data from the Boston Area Research Initiative (BARI), hosted on Harvard Dataverse. BARI curates administrative, commercial, and social media data about Boston neighborhoods at multiple geographic scales.

Think of tidybari as "tidycensus for BARI data" — it lets you download neighborhood-level data with a single function call, optionally including spatial geometries for mapping.

## Installation

You can install tidybari from GitHub:

```{r install}
# install.packages("remotes")
remotes::install_github("joshrrosen/tidybari")
```

## Loading the Package

```{r setup, eval = TRUE}
library(tidybari)

# For spatial analysis and mapping
library(sf)
library(ggplot2)
library(dplyr)
```

## Discovering Available Data

### Browse the Dataset Catalog

The `bari_catalog()` function shows all available datasets:

```{r catalog, eval = TRUE}
# View all datasets
catalog <- bari_catalog()
head(catalog)

# Filter to a specific domain
permits_catalog <- bari_catalog(domain = "permits")
permits_catalog

# Filter by year
catalog_2020 <- bari_catalog(year = 2020)
```

The catalog shows:
- **domain**: Data source (911, permits, airbnb, etc.)
- **level**: Geographic/data level (records, ecometrics_ct, etc.)
- **year**: Data year
- **doi**: Harvard Dataverse persistent ID
- **filename**: Exact filename on Dataverse

### Explore Available Variables

Use `load_variables()` to see what ecometric measures are available:

```{r variables, eval = TRUE}
# View all variables
all_vars <- load_variables()
all_vars

# Variables for a specific domain
permits_vars <- load_variables("permits")
permits_vars

# In RStudio, use View() for interactive browsing
# View(load_variables())
```

## Downloading Ecometric Data

Ecometrics are neighborhood-level indices that measure social and physical characteristics of place. They're derived from administrative records but aggregated to protect privacy.

### Basic Usage

```{r ecometrics-basic}
# Get 911 call ecometrics (longitudinal data covering 2010-2022)
# Note: 911 ecometrics are only available as longitudinal data
data_911 <- get_ecometrics(
  domain = "911",
  longitudinal = TRUE
)

head(data_911)
```

### Select Specific Variables

```{r ecometrics-variables}
# Get only investment and growth measures from permits
permits <- get_ecometrics(
  domain = "permits",
  geography = "tract",
  year = 2020,
  variables = c("investment", "growth")
)

head(permits)
```

### Different Geographic Levels

```{r ecometrics-geography}
# Census tract level (default)
tract_data <- get_ecometrics("permits", year = 2020)

# Census block group level (finer detail)
bg_data <- get_ecometrics(
  "permits",
  geography = "block_group",
  year = 2020
)

# Land parcel level (property-level, permits only)
parcel_data <- get_ecometrics(
  "permits",
  geography = "parcel",
  year = 2020
)
```

### Longitudinal Data

```{r ecometrics-longitudinal}
# Get data across multiple years
permits_long <- get_ecometrics(
  "permits",
  longitudinal = TRUE
)

# This returns data for all available years
head(permits_long)
```

## Downloading Record-Level Data

Some datasets provide individual records (like permit applications or Airbnb listings):

```{r records}
# Get building permit applications (most recent year available)
permits_records <- get_records("permits")
head(permits_records)

# Airbnb listings (not year-specific)
airbnb <- get_records("airbnb")

# Yelp reviews
yelp <- get_records("yelp")
```

**Note:** Raw 911 dispatch records are restricted by BARI for privacy. Only aggregated ecometrics are publicly available.

```{r records-911, error = TRUE, eval = TRUE}
# This will error with a helpful message
try(get_records("911"))
```

## Spatial Analysis and Mapping

### Get Data with Geometries

The most powerful feature of tidybari is the ability to get data with spatial geometries ready for mapping:

```{r spatial-data}
# Get ecometrics with geometries
permits_spatial <- get_ecometrics(
  domain = "permits",
  geography = "tract",
  year = 2020,
  geometry = TRUE  # This adds spatial boundaries
)

# Result is an sf object
class(permits_spatial)
```

### Map Ecometrics

```{r map-basic}
# Map investment ecometric
ggplot(permits_spatial) +
  geom_sf(aes(fill = investment)) +
  scale_fill_viridis_c(option = "plasma") +
  theme_minimal() +
  labs(
    title = "Building Permit Investment by Census Tract",
    subtitle = "Boston, 2020",
    fill = "Investment\nIndex"
  )
```

### Get Geography Boundaries Separately

You can also get spatial boundaries without data:

```{r geography, eval = TRUE}
# Get census tract boundaries
tracts <- get_geography("tract")
plot(tracts["NAME"])

# Get block group boundaries
bg <- get_geography("block_group")

# Get BPDA neighborhood boundaries
neighborhoods <- get_geography("neighborhood")
```

### Custom Maps

```{r map-custom}
# Get data
permits <- get_ecometrics(
  "permits",
  year = 2020,
  geometry = TRUE
)

# Create a custom map
ggplot(permits) +
  geom_sf(aes(fill = growth), color = "white", size = 0.1) +
  scale_fill_gradient2(
    low = "blue",
    mid = "white",
    high = "red",
    midpoint = median(permits$growth, na.rm = TRUE)
  ) +
  theme_void() +
  labs(
    title = "Urban Growth Patterns",
    subtitle = "Derived from building permits, 2020",
    fill = "Growth Index",
    caption = "Data: Boston Area Research Initiative"
  )
```

## Caching for Performance

Downloaded files are automatically cached locally to improve performance:

```{r cache, eval = TRUE}
# View cache directory
bari_cache_dir()

# Set a custom cache location
bari_cache_dir("~/my_bari_cache")

# Clear the cache (free up disk space)
bari_cache_clear()
```

Control caching behavior in download functions:

```{r cache-control}
# Force fresh download even if cached
data <- get_ecometrics("permits", year = 2020, refresh = TRUE)

# Disable caching for this download
data <- get_ecometrics("permits", year = 2020, cache = FALSE)
```

## Combining Multiple Datasets

```{r combine}
# Get 911 longitudinal ecometrics and permits data for a common year
data_911 <- get_ecometrics("911", longitudinal = TRUE) %>%
  dplyr::filter(year == 2020)
data_permits <- get_ecometrics("permits", year = 2020)

# Join by census tract ID
combined <- data_911 %>%
  inner_join(data_permits, by = "CT_ID_10", suffix = c("_911", "_permits"))

# Now you can analyze relationships
library(ggplot2)
ggplot(combined, aes(x = crime, y = investment)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm") +
  theme_minimal() +
  labs(
    title = "Crime and Investment Patterns",
    x = "Crime Index (911 calls)",
    y = "Investment Index (permits)"
  )
```

## Example Workflow

Here's a complete workflow from discovery to visualization:

```{r workflow}
# 1. Discover what's available
load_variables("permits")

# 2. Get data with geometries
data <- get_ecometrics(
  domain = "permits",
  geography = "tract",
  year = 2020,
  variables = c("investment", "growth"),
  geometry = TRUE
)

# 3. Quick exploration
summary(data)

# 4. Create a map
ggplot(data) +
  geom_sf(aes(fill = investment)) +
  scale_fill_viridis_c() +
  theme_minimal() +
  labs(title = "Investment Patterns in Boston")

# 5. Save for later use
# sf::st_write(data, "boston_permits_2020.geojson")
```

## Data Citation

All data come from the Boston Area Research Initiative. When using BARI data in publications, please cite appropriately:

```{r citation, eval = TRUE}
# Find the DOI for your dataset
catalog <- bari_catalog(domain = "permits", year = 2020)
catalog$doi[1]
```

**Example citation:**

> Boston Area Research Initiative, 2020. Building Permits. Harvard Dataverse. https://doi.org/10.7910/DVN/B7DHBK

## Getting Help

- **Function documentation**: `?get_ecometrics`, `?bari_catalog`, etc.
- **Package website**: [https://github.com/joshrrosen/tidybari](https://github.com/joshrrosen/tidybari)
- **BARI website**: [https://cssh.northeastern.edu/bari/](https://cssh.northeastern.edu/bari/)
- **Report issues**: [https://github.com/joshrrosen/tidybari/issues](https://github.com/joshrrosen/tidybari/issues)
- **BARI data questions**: [[email protected]](mailto:[email protected])

## Additional Resources

- [BARI Data Portal on Harvard Dataverse](https://dataverse.harvard.edu/dataverse/BARI)
- [tidycensus package](https://walker-data.com/tidycensus/) (similar API for US Census data)
- [Simple Features for R](https://r-spatial.github.io/sf/) (sf package documentation)

---

**Happy mapping!** 🗺️
